---
title: "*R-code*"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Required packages

```{r}
library(magrittr) # tidy coding
library(rredlist) # IUCN Queries
library(readr) # reading tables
library(taxize) # checking taxon names
```


## Query IUCN data

The [IUCN Redlist API](http://apiv3.iucnredlist.org/api/v3/docs) is queried using the `rredlist`[^1] package. Using this service requires a token (see Link above on how to get access). The function `iucn_query` executes the query and outputs four files that are saved within the folder `data`.

```{r}
## load the token. The token is not provided here!
load("RedListAPI.RData")
```

```{r, eval=FALSE}
## read taxon names
taxon <- readxl::read_xlsx("data/dataset_combined.xlsx")
taxon <- taxon[["taxon.iucn"]] %>% as.character()

## source script and run the query
source("R/iucn_query.R")
iucn_query(taxon = taxon, key = RedLIstAPI)
```

```{r, eval=FALSE, echo=FALSE}
## check for problems to access the IUCN data
malavi <- readr::read_delim("data/MalAvi_host_and_site_coord.csv", delim = ";")
load("data/unknown.names.RData")
```

### Extract habitats 

For most species there a several habitats listed (from a total of > 90 terms), sometimes distinguished by `season` (e.g. Breeding Season, Non-Breeding Season, Passage & Resident). Additionally, a boolean codes `majorimportance` im some cases. Therefore, as a mean of selecting main habitats in a objective manner, we selected natural habitats (i.e. ignoring all terms including the keyword "Artifical") for each species as follows:

1. When available, we only considered habitats listed as `majorimportance`
2. When seasons were distinguished we only selected `Breeding Season`
3. Habitat with *'Marginal'* `suitability` were not considered.  

```{r, warning=F, message=F}
## load output of iucn_query
load("data/habitats.RData") 
load("data/common.names.RData")
df <- readxl::read_xlsx("data/dataset_combined.xlsx") %>% 
  dplyr::filter(., taxon.iucn %in% names(habitats))

## select habitats for each species
source("R/select_habitats.R")
habitat.matrix <- lapply(df[["taxon.iucn"]], select_habitats, habitats = habitats) %>%   do.call("rbind",.)  %>% 
  set_rownames(., df[["taxon.iucn"]])

## get all terms 
habitat.names <-
  lapply(df[["taxon.iucn"]], select_habitats, habitats = habitats, out = "names") %>% 
  set_names(., df[["taxon.iucn"]]) %>% 
  unlist()
habitat.names <- data.frame(taxon.iucn = names(habitat.names),
                            habitats = habitat.names)

names(common.names) <- c("taxon.iucn", "taxon.english")


## get coarser categories, ie forest, grassland, wetland etc.
source("R/habitat_categorisation.R")
habitat.coarse <- habiat_categorisation(habitat.matrix = habitat.matrix, 
                              path = "data/simplified_habitat_scheme.xlsx" )


df[["Common name"]] <- NULL
df[["Main Habitat Breeding season"]] <- NULL
```

## EltonTraits

Standardised Species-level foraging and diet attributes are available from the [EltonTraits 1.0](http://www.esapubs.org/archive/ecol/E095/178/)[^2] database. In this dataset, foraging strata and diet types (*'Elton´s Niche'*) are available on a semi-quantitative scale giving the relative importance for each taxon (values from 0 to 100 in steps of 10). 

```{r, eval=FALSE}
## download EltonTraits
download.file(url = "http://www.esapubs.org/archive/ecol/E095/178/BirdFuncDat.txt",
              destfile = "data/BirdFuncDat.txt",
              quiet = T,
              cacheOK = T)
```

```{r, warning=FALSE, message=FALSE}
## fewer species than in dataset due to lumping/splitting of species
BirdFuncDat <-
  read_delim("data/BirdFuncDat.txt", "\t", trim_ws = T, skip_empty_rows = T) 
# %>% 
#   dplyr::filter(., Scientific %in% df[["taxon.eltontraits"]])
BirdFuncDat[["taxon.eltontraits"]] <- BirdFuncDat[["Scientific"]]


df2 <- df
df2 <- lapply(1:nrow(df2), function(x) {
  if (is.na(df2[["taxon.hbw"]][x])) df2[["taxon.hbw"]][x] <- df2[["taxon.iucn"]][x]
return(df2[x,])
}) %>% 
  do.call("rbind",.)

## Function to find synonyms in species names
source("R/match_synonyms.R")
dfx <- data.frame(taxon.iucn = df2[["taxon.iucn"]],
                 taxon.hbw = df2[["taxon.hbw"]],
                 taxon.malavi = df2[["taxon.malavi"]],
                 taxon.english = common.names[["taxon.english"]]) %>%
  dplyr::mutate_all(., as.character)

dfx[["taxon.elton"]] <- 
match_synonyms(df = dfx,
BirdFuncDat = BirdFuncDat)

View(dplyr::filter(dfx, is.na(taxon.elton)))
write.csv(data.frame(taxon = dfx$taxon.iucn, match = dfx$taxon.elton), "data/temp.csv")
```

## AnAge 

The [AnAge](https://genomics.senescence.info/species/)[^3] Database provides proxies of longevity

```{r, eval=FALSE}
## download zp file
download.file(url = "http://genomics.senescence.info/species/dataset.zip",
              destfile = "data/AnAge.zip",
              quiet = T)
## extract data table
unzip(zipfile = "data/AnAge.zip", files = "anage_data.txt", exdir = "data")
## delete zip file
unlink("data/AnAge.zip")

AnAge <- 
  read_delim("data/anage_data.txt", "\t", trim_ws = T, skip_empty_rows = T)
AnAge[["taxon.iucn"]] <- paste(AnAge[["Genus"]], AnAge[["Species"]]) 
```


## Merge data

* Get rid of captive individuals (`r (275/7248)*100`)
* Get rid of hybrides 
* Define climate zones based on climate lcoation

Merge all datasets into one and format. 

```{r, message=FALSE, warning=FALSE}
## Join datasets
dataset <- 
  ## common name
  dplyr::left_join(common.names, df, by = "taxon.iucn") %>% 
  ## habitat names
  dplyr::left_join(., habitat.names, by = "taxon.iucn") %>% 
  ## habitat categories
  dplyr::left_join(., habitat.coarse, by = "taxon.iucn") %>% 
  ## Elton´s niche
  dplyr::left_join(., BirdFuncDat, by = "taxon.eltontraits")
  
## load IUCN classifications
load("data/redlist.RData")
redlist[["taxon.iucn"]] <- redlist[["scientific_name"]]
dataset <- dplyr::left_join(dataset, redlist, by = "taxon.iucn")

## reformat
data <- data.frame(
  ## Taxonomy
  scientific_name = as.factor(dataset[["taxon.iucn"]]),
  scientfic_malavi = as.factor(dataset[["taxon.malavi"]]),
  order = as.factor(dataset[["order"]]),
  family = as.factor(dataset[["family"]]),
  genus = as.factor(dataset[["genus"]]),
  pass_nonpass = as.factor(dataset[["PassNonPass"]]),
  common_name = as.factor(dataset[["main_common_name"]]),
  ## Redlist category
  redlist_category = as.factor(dataset[["category"]]),
  population_trend = as.factor(dataset[["population_trend"]]),
  movements = as.factor(dataset[["Movements"]]),
  ## distribution
  elevation_upper = as.numeric(dataset[["elevation_upper"]]),
  elevation_lower = as.numeric(dataset[["elevation_lower"]]),
  ## diet & foraging
  nocturnal = as.factor(dataset[["Nocturnal"]]),
  diet_category = as.factor(dataset[["Diet-5Cat"]]), 
  body_mass = dataset[["BodyMass-Value"]],
  forstrat_watbelowsurf = dataset[["ForStrat-watbelowsurf"]],
  forstrat_wataroundsurf = dataset[["ForStrat-wataroundsurf"]],
  forstrat_ground = dataset[["ForStrat-ground"]],
  forstrat_understory = dataset[["ForStrat-understory"]],
  forstrat_midhigh = dataset[["ForStrat-midhigh"]],
  forstrat_canopy = dataset[["ForStrat-canopy"]],
  forstrat_aearial = dataset[["ForStrat-aerial"]],
  ## habitat and nest sites 
  nest_type = as.factor(dataset[["Nest type"]]),
  nest_site = as.factor(dataset[["Nest site"]]),
  habitat_names = as.character(dataset[["habitats"]]),
  habitat_category = as.factor(dataset[["habitat"]])
  )
```


```{r}
x <- dplyr::filter(dataset, 
                   genus %in% c("Turdus", "Otus", "Buteo", "Falco", "Parus",
                                "Passer", "Aquila", "Dendrocopus",
                                "Asio"))
rownames(x) <- x$taxon.iucn
data.frame(var = names(x), seq = 1:ncol(x))

x <- x[,c(22:31, 36:42, 47, 12, 6:7)]

hc <- hclust(dist(x), "ave")
plot(hc)
plot(hc, hang = -1)
```


```{r}
```


Bring together all the resources ... and:

1. Unifying habitats: (e.g. merging related categories & building hierachical categories)
2. Adding additional RedList variables
4. Adding sample metadata (e.g. coords)
5. Solve taxnonmy issues by two columns 'taxon.malavi' and 'taxon.hbw' allowing to match entries in a two-way join.

* Having a binomial coding single/major importance habitat?

[^1]: Scott Chamberlain (2018). rredlist: 'IUCN' Red List Client. R package version 0.5.0. https://CRAN.R-project.org/package=rredlist
[^2]: Hamish Wilman, Jonathan Belmaker, Jennifer Simpson, Carolina de la Rosa, Marcelo M. Rivadeneira, and Walter Jetz. 2014. EltonTraits 1.0: Species-level foraging attributes of the world’s birds and mammals. Ecology 95:2027. http://dx.doi.org/10.1890/13-1917.1
[^3]: Tacutu, R., Craig, T., Budovsky, A., Wuttke, D., Lehmann, G., Taranukha, D., Costa, J., Fraifeld, V. E., de Magalhaes, J. P. (2013) "Human Ageing Genomic Resources: Integrated databases and tools for the biology and genetics of ageing." Nucleic Acids Research 41(D1):D1027-D1033