---
title: "*R-code*"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Required packages

```{r}
library(magrittr) # tidy coding
library(rredlist) # IUCN Queries
library(readr) # reading tables
```


## Query IUCN data

The [IUCN Redlist API](http://apiv3.iucnredlist.org/api/v3/docs) is queried using the `rredlist`[^1] package. Using this service requires a token (see Link above on how to get access). The function `iucn_query` executes the query and outputs four files that are saved within the folder `data`.

```{r}
## load the token. The token is not provided here!
load("RedListAPI.RData")
```

```{r, eval=T}
## read taxon names
taxon <- readxl::read_xlsx("data/dataset_combined.xlsx")
taxon <- taxon[["taxon.iucn"]] %>% as.character() 
# taxon <- taxon[1049:1576]

## source script and run the query
source("R/iucn_query.R")
iucn_query(taxon = taxon, key = RedLIstAPI, .update = T)
```

```{r, eval=T, echo=FALSE}
## check for problems to access the IUCN data
malavi <- readr::read_delim("data/MalAvi_host_and_site_coord.csv", delim = ";")
load("data/unknown.names.RData")
```

### Extract habitats 

For most species there a several habitats listed (from a total of > 90 terms), sometimes distinguished by `season` (e.g. Breeding Season, Non-Breeding Season, Passage & Resident). Additionally, a boolean codes `majorimportance` im some cases. Therefore, as a mean of selecting main habitats in a objective manner, we selected natural habitats (i.e. ignoring all terms including the keyword "Artifical") for each species as follows:

1. When available, we only considered habitats listed as `majorimportance`
2. When seasons were distinguished we only selected `Breeding Season`
3. Habitats of *'Marginal'* `suitability` were not considered.  

```{r, warning=F, message=F}
## load output of iucn_query
load("data/habitats.RData") 
load("data/common.names.RData")
df <- readxl::read_xlsx("data/dataset_combined.xlsx") 

## select habitats for each species
source("R/select_habitats.R")
habitat.matrix.major <- pbapply::pblapply(df[["taxon.iucn"]], select_habitats,
                               habitats = habitats,
                               major.importance = T) %>%
  do.call("rbind",.)  %>% 
  set_rownames(., df[["taxon.iucn"]])

habitat.matrix.all <- pbapply::pblapply(df[["taxon.iucn"]], select_habitats,
                               habitats = habitats,
                               major.importance = F) %>%
  do.call("rbind",.)  %>% 
  set_rownames(., df[["taxon.iucn"]])

## get all terms 
habitat.names <-
  pbapply::pblapply(df[["taxon.iucn"]], select_habitats, habitats = habitats, out = "names") %>% 
  set_names(., df[["taxon.iucn"]]) %>% 
  unlist()

habitat.names <- data.frame(taxon.iucn = names(habitat.names),
                            habitats = habitat.names)

names(common.names) <- c("taxon.iucn", "taxon.english")


## get coarser categories, ie forest, grassland, wetland etc.
source("R/habitat_categorisation.R")
habitat.coarse <- habiat_categorisation(habitat.matrix = habitat.matrix.major, 
                              path = "data/simplified_habitat_scheme.xlsx" )

df[["taxon.common"]] <- NULL
df[["habitats"]] <- NULL
```

## EltonTraits

Standardised Species-level foraging and diet attributes are available from the [EltonTraits 1.0](http://www.esapubs.org/archive/ecol/E095/178/)[^2] database. In this dataset, foraging strata and diet types (*'Elton´s Niche'*) are available on a semi-quantitative scale giving the relative importance for each taxon (values from 0 to 100 in steps of 10). 

```{r, eval=FALSE}
## download EltonTraits
download.file(url = "http://www.esapubs.org/archive/ecol/E095/178/BirdFuncDat.txt",
              destfile = "data/BirdFuncDat.txt",
              quiet = T,
              cacheOK = T)
```

```{r, warning=FALSE, message=FALSE}
## load all entries that are available
BirdFuncDat <-
  read_delim("data/BirdFuncDat.txt", "\t", trim_ws = T, skip_empty_rows = T) 
BirdFuncDat[["taxon.eltontraits"]] <- BirdFuncDat[["Scientific"]]

## if there is no taxon hbw entered, copy iucn nomencalture as dummy variable
df2 <- df
df2 <- lapply(1:nrow(df2), function(x) {
  if (is.na(df2[["taxon.hbw"]][x])) df2[["taxon.hbw"]][x] <- df2[["taxon.iucn"]][x]
return(df2[x,])
}) %>% 
  do.call("rbind",.)

## Function to find synonyms in species names to match entries
source("R/match_synonyms.R")
dfx <- data.frame(taxon.iucn = df2[["taxon.iucn"]],
                  taxon.hbw = df2[["taxon.hbw"]],
                  taxon.malavi = df2[["taxon.malavi"]],
                  taxon.elton = df2[["taxon.eltontraits"]]) %>%
  dplyr::left_join(., common.names, by = "taxon.iucn") %>% 
  dplyr::mutate_all(., as.character)

dfx[["taxon.elton"]] <- match_synonyms(df = dfx,
                                       BirdFuncDat = BirdFuncDat)

#View(dplyr::filter(dfx, is.na(taxon.elton)))
write.csv(data.frame(taxon = dfx$taxon.iucn, match = dfx$taxon.elton), "data/temp.csv")
```

## AnAge 

The [AnAge](https://genomics.senescence.info/species/)[^3] Database provides proxies of longevity

```{r, eval=FALSE}
## download zp file
download.file(url = "http://genomics.senescence.info/species/dataset.zip",
              destfile = "data/AnAge.zip",
              quiet = T)
## extract data table
unzip(zipfile = "data/AnAge.zip", files = "anage_data.txt", exdir = "data")
## delete zip file
unlink("data/AnAge.zip")

AnAge <- 
  read_delim("data/anage_data.txt", "\t", trim_ws = T, skip_empty_rows = T)
AnAge[["taxon.iucn"]] <- paste(AnAge[["Genus"]], AnAge[["Species"]]) 
```


## Merge data

Taxonomic concepts are not always in concordance between authorities such as *HBW*, *BirdLife*, *Clements* or *IOC*. Furthermore, there are constant changes due to lumping and splitting of species but also renaming for other reasons occurs. Hence, we had to match entries stemming from (i) **MalAvi** (*which date?*), (ii) **HBW Alive** (December 2019) and (iii) **EltonTraits** (*2014?*). Notably, information for species which were until recently considered as subspecies only (e.g. Oriental Reed-warbler *Acrocephalus orientalis* & Great Reed-warbler *A. arundinaceus*), were not available from the EltonTraits database for considerable number of taxons. Here, we used the data available for respective sister species. Due to the high similarity and uncertain in distinguishin between species and subspecies these data are arguably represensative of the ecology of both taxons. (**Rephrasing needed !!!**) 

We have duplicates!!!

Merge all datasets into one and format and simplify some factor levels. 

* How to deal with captive individuals?
* How to deal with hybrides ?

```{r, message=FALSE, warning=FALSE}
## Join datasets
dataset <- 
  ## common name
  dplyr::left_join(df, unique.data.frame(common.names), by = "taxon.iucn") %>% 
  ## habitat names
  dplyr::left_join(., unique.data.frame(habitat.names), by = "taxon.iucn") %>% 
  ## habitat categories
  dplyr::left_join(., unique.data.frame(habitat.coarse), by = "taxon.iucn") %>% 
  ## Elton´s niche
  dplyr::left_join(., BirdFuncDat, by = "taxon.eltontraits")
  
## load IUCN classifications
load("data/redlist.RData")
redlist[["taxon.iucn"]] <- redlist[["scientific_name"]]
dataset <- dplyr::left_join(dataset, unique.data.frame(redlist), by = "taxon.iucn")

## reformat
data <- data.frame(
  ## Taxonomy
  scientific_name = as.factor(dataset[["taxon.iucn"]]),
  scientfic_malavi = as.factor(dataset[["taxon.malavi"]]),
  order = as.factor(dataset[["order"]]),
  family = as.factor(dataset[["family"]]),
  genus = as.factor(dataset[["genus"]]),
  pass_nonpass = as.factor(dataset[["PassNonPass"]]),
  common_name = as.factor(dataset[["main_common_name"]]),
  ## Redlist category
  redlist_category = as.factor(dataset[["category"]]),
  population_trend = as.factor(dataset[["population_trend"]]),
  movements = as.factor(dataset[["movements"]]),
  ## distribution
  elevation_upper = as.numeric(dataset[["elevation_upper"]]),
  elevation_lower = as.numeric(dataset[["elevation_lower"]]),
  ## diet & foraging
  nocturnal = as.factor(dataset[["Nocturnal"]]),
  diet_category = as.factor(dataset[["Diet-5Cat"]]), 
  body_mass = dataset[["BodyMass-Value"]],
  forstrat_watbelowsurf = dataset[["ForStrat-watbelowsurf"]],
  forstrat_wataroundsurf = dataset[["ForStrat-wataroundsurf"]],
  forstrat_ground = dataset[["ForStrat-ground"]],
  forstrat_understory = dataset[["ForStrat-understory"]],
  forstrat_midhigh = dataset[["ForStrat-midhigh"]],
  forstrat_canopy = dataset[["ForStrat-canopy"]],
  forstrat_aearial = dataset[["ForStrat-aerial"]],
  ## habitat and nest sites 
  nest_type = as.factor(dataset[["nest.type"]]),
  nest_site = as.factor(dataset[["nest.site"]]),
  habitat_names = as.character(dataset[["habitats"]]),
  habitat_category = as.factor(dataset[["habitat"]])
  )
```


```{r}

data$diet_category %>% summary()
  data$movements %>% summary()

data$movements[data$movements == "migratory"] %>% length()


  data$nest_type %>% summary()
  data$nest_site %>% summary()
  data$habitat_category %>% summary()

write.csv(data, "data/dataset_combined_final.csv")  

habitat.data.major <- cbind(data[,1:7], habitat.matrix.major) %>% 
  as.data.frame()
write.csv(habitat.data.major, "data/habitat_matrix_majorimportance.csv")
habitat.data.all <- cbind(data[,1:7], habitat.matrix.all) %>% 
  as.data.frame()
write.csv(habitat.data.all, "data/habitat_matrix.csv")
```

```{r}
## some checks
if (any(is.na(data[["redlist_category"]]))) stop("NA´s in IUCN data")
if (any(is.na(data[["diet_category"]]))) stop("NA´s in elton traits")
```

## Trying stuff

```{r, eval=FALSE}
x <- dplyr::filter(dataset, 
                   genus %in% c("Turdus", "Otus", "Buteo", "Falco", "Parus",
                                "Passer", "Aquila", "Dendrocopus",
                                "Asio")) %>% 
  as.data.frame()

rownames(x) <- x$taxon.malavi
data.frame(var = names(x), seq = 1:ncol(x))

x <- x[,c(22:31, 36:42, 47, 12, 6:7)]

hc <- hclust(dist(x), "ave")
plot(hc)
plot(hc, hang = -1)
```

```{r, eval=FALSE}
x <- dplyr::filter(habitat.data.all, 
                   genus %in% c("Turdus", "Otus", "Buteo", "Falco", "Parus",
                                "Passer", "Aquila", "Dendrocopus",
                                "Asio")) %>% 
  unique.data.frame()
x <- habitat.data.all %>% unique.data.frame()
rownames(x) <- x$scientfic_malavi
data.frame(var = names(x), seq = 1:ncol(x))

x <- x[,8:80]

hc <- hclust(dist(x), "ave")
plot(hc)
plot(hc, hang = -1)
```

Bring together all the resources ... and:

1. Unifying habitats: (e.g. merging related categories & building hierachical categories)
2. Adding additional RedList variables
4. Adding sample metadata (e.g. coords)
5. Solve taxnonmy issues by two columns 'taxon.malavi' and 'taxon.hbw' allowing to match entries in a two-way join.

* Having a binomial coding single/major importance habitat?

[^1]: Scott Chamberlain (2018). rredlist: 'IUCN' Red List Client. R package version 0.5.0. https://CRAN.R-project.org/package=rredlist
[^2]: Hamish Wilman, Jonathan Belmaker, Jennifer Simpson, Carolina de la Rosa, Marcelo M. Rivadeneira, and Walter Jetz. 2014. EltonTraits 1.0: Species-level foraging attributes of the world’s birds and mammals. Ecology 95:2027. http://dx.doi.org/10.1890/13-1917.1
[^3]: Tacutu, R., Craig, T., Budovsky, A., Wuttke, D., Lehmann, G., Taranukha, D., Costa, J., Fraifeld, V. E., de Magalhaes, J. P. (2013) "Human Ageing Genomic Resources: Integrated databases and tools for the biology and genetics of ageing." Nucleic Acids Research 41(D1):D1027-D1033